prolfqua - An R-package for Analysing Proteomics Label Free Quantification Experiments

### 2022-03-18 inst/MyArticle/paper.Rmd
Homogenize math notations
- abstract
- math notation,e.g., eqnarray t-statistics, p-statistics

---
title: prolfqua - R-package for Analysing Proteomics Label Free Quantification Experiments
author: 
  - name: Witold Wolski
    affil: 1; 2
    orcid: '0000-0002-6468-120X'
  - name: Paolo Nanni
    affil: 1
    orcid: '0000-0001-8429-3557'
  - name: Jonas Grossmann
    affil: 1; 2
    orcid: '0000-0002-6899-9020'
  - name: Maria D'Errico
    affil: 1; 2
    orcid: '0000-0001-9467-9058'
  - name: Ralph Schlapbach
    affil: 1
    orcid: '0000-0002-7488-4262'
  - name: Christian Panse
    affil: 1; 2
    orcid: '0000-0003-1975-3064'
affiliation:
  - num: 1
    address: Functional Genomics Center Zurich (https://www.fgcz.ch/)
  - num: 2
    address: Swiss Institute of Bioinformatics (https://www.sib.swiss/)
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 7, fig.height = 5)

See ACS guidelines [https://publish.acs.org/publish/author_guidelines?coden=ancham#manuscript_text_components]

# Abstract (80–250 words)

Protein mass spectrometry is widely used for quantitative proteomics studies and in particular for relative protein quantification. Nevertheless, there is still a need for a flexible and easy-to-use, transparently supporting a variety of well principled statistical inference procedures data analysis application programmer interface in R. 
The _prolfqua_ R package integrates essential steps of the mass spectrometry-based label-free quantification data analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. We use _prolfqua_ to visualize and model simple experimental designs with a single explanatory variable and complex experiments with multiple factors. Furthermore, we use _prolfqua_ to benchmark data acquisition, data preprocessing or data modeling methods. Furthermore, we use _prolfqua_ to develop highly customizable, visually appealing, and interactive data analysis reports in pdf or HTML format. The R package is available from https://github.com/fgcz/prolfqua.

@@

Label-Free Quantification (LFQ) Proteomics Experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment vs. control are used [TODO: cite]. More recently, more complex experimental designs are implemented with increasing numbers of samples, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [TODO: cite]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a factor or interactions are significant, or differences between factor levels can be estimated, and their significance tested.  Proteins can be ordered or grouped using the estimated fold-change, $t$-statistics, scaled $p$-value, or false discovery rate (FDR) and then subjected to gene set enrichment or over-representation analysis [TODO ...] or validation experiments.

@@

Models in `r BiocStyle::Githubpkg("prolfqua")` are specified using R’s linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure [@faraway2016extending,@MASS2002] is an advantage when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis which was performed with `r BiocStyle::Githubpkg("prolfqua")` using base R or other programming languages by reading the analysis script. However, acknowledging the R formula interface’s complexity to non-statisticians and the popularity of `r BiocStyle::Biocpkg("MSstats")`, we provide the functionality to suggest a model formula from the sample annotation provided in tabular form similarily to `r BiocStyle::Biocpkg("MSstats")` [@Choi2014].

@@

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented. This makes it easy to reproduce an analysis performed with _prolfqua_ in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and p-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We use R6 to model the statistical modeling functionality in `prolfqua` (see Figure \@ref(fig:ContrastUML)).

(ref:ContrastUML) UML diagram of modeling and contrast related classes. Different strategies e.g. _lm_, _lmer_, and _glm_ can be used to fit various types of models using a model builder (see code). All classes computing contrasts implement the _Contrast_ interface.

@@

$$s_p^2=\frac{\sum_{i=1}^k (n_i - 1)s_i^2}{\sum_{i=1}^k (n_i - 1)},$$

@@

$$
s = \sqrt{\frac{2n_g s_p^2}{n}},
$$

@@

$\tilde{t}_{pj} = \frac{t_{pj} s_p}{\tilde{s}_p}$

@@

To summarize peptide level models to protein models we did applied the method suggested by [@Suomi2017bEnhanced], that is to use the median scaled p-value of the peptide models and cumulative distribution function of the beta distribution function to determine a regulation probability of the protein.

To obtain the median p-value of the protein, we first rescaled the peptide p-values by taking the direction of the fold-change $\hat \beta$ into account, i.e.:

@@

$$
\tilde{p} = 1 - |\tilde{p}_{s}|
$$
Because we used the median as the i-th order statistic  $i = n/2 +0.5$. Therefore, $\gamma = i = n/2 + 0.5$ and $\delta = n - i + 1 = n - (n/2 + 0.5) + 1 = n/2 + 0.5 = \gamma$ are used to parameterize the CDF of the Beta distribution. We implemented this method in the class `ContrastROPECA` (Figure \@ref(fig:ContrastUML)).

@@

(ref:exampleContrasts) Left panel - histogram of p-values. Right panel - volcano plots.

@@

The Benchmark functionality of _prolfqua_ includes ROC curves and computes partial areas under those curves (pAUC) and other scores, e.g. the false discovery proportion FDP. We use it to study how well linear, mixed effect models or p-value moderation model quantitative mass spectrometric high throughput experiments and compare it with results produced by MSstats [@MSstats2014] and by proDA [@bioxrvproDA2020]. Figure \@ref(fig:FDRfdp)

@@

We can use various statistics to examine the alternative hypothesis $\beta \ne 0$: the contrast estimate, i.e. the $\log_2$ fold-change $\beta$, the $t$-statistic $\frac{\beta}{\sqrt{var(\beta)}}$, or the $p-value$ and moderated $p-value$. For each statistic and each value of the statistics we then compute a confusion matrix (see Table \@ref(tab:confusionMatrix)). From the confusion matrix we obtain measures such as true positive rate (TPR), false positive rate (FPR), or false discovery proportion (FDP) which are given by:

@@

$$
\begin{aligned}
TPR &= \frac{TP}{TP+FN} = \frac{TP}{P}\\
FPR &= \frac{FP}{FP+TN} = \frac{FP}{N}\\
FDP &= \frac{FP}{TP + FP} = \frac{FP}{R}
\end{aligned}
$$

@@

Figure \@ref(fig:FDRfdp) left panel, shows how various estimates obtained from the models, i.e. $\log_2(FC)$, $t-statistics$ and scaled p-values allow to identify true positives (TP) given a false positive rate. When testing for differential expression, the $\log_2(FC)$ is outperformed by than the $t$-statistic or $p$-value. 
If an accurate estimate of the $log_2(FC)$ is essential, the linear models fitted to protein intensities, estimated using Tukey's median polish, show the best performance. We speculate that this is because of the robustness of Tukey's median polish method. This hypothesis could be examined, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`, into the benchmark.

There are only minor differences in the p-AUC when using the t-statistics or the scaled p-value  (see Figure \@ref(fig:FDRfdp) right panel). In order to compute the confusion matrices based on the p-value we first need to rescale it (see Equation \@ref(eq:scalepvalue)). Thus, the p-value is a function of the t-statistics and the degrees of freedom. Interestingly, the pAUC increases when using p-values instead of the t-statistics for linear models, while it decreases for mixed effect models. The reason is erroneous denominator degrees of freedom estimation for many proteins in the mixed-effect models.

@@

(ref:FDRfdp) Left panel - Partial area under the ROC curve at $10\%$ FPR for all contrasts and three different statistics: the difference among conditions, the scaled p-value (sign(diff) * p.value) and the t-statics (higher is better). The red line average area under the curve. Left panel - Number and percentage of estimated contrasts for each modeling method (higher is better). Right panel - Compare FDR estimate with false discovery proportion (FDP) closer to the diagnoal (red continous line is better).

@@

The probabilistic dropout analysis implemented in the _proDA_ produces inferences comparable to other methods (Figure \@ref(fig:FDRfdp) for the largest number of proteins (Figure \@ref(fig:FDRfdp) left panel). Moreover, the estimated fold changes are unbiased and show high diagnostic accuracy. Furthermore, the scaled p-values and t-statics are on par with those obtained by the linear model with variance moderation. Therefore, we plan to integrate the _proDA_ package as an additional modeling option into _prolfqua_ (see Figure \@ref(fig:ContrastUML)).

Leveraging on these computer experiments, we can provide the following advice. First, aggregate the peptide intensities to protein intensities using a robust regression method. Second, use linear models as long as they suffice to model your experimental design. Use mixed effect models only if you have to (e.g., repeated measurements). If you use fixed-effect linear models, apply variance moderation since this improves the t-statistics and p-value estimates. If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the fold change. You can select proteins to study them further based on the FDR, although expect a $bias$.

@@

>>>>

---
title: prolfqua - An R-package for Analysing Proteomics Label Free Quantification Experiments
author: 
  - name: Witold Wolski
    affil: 1; 2
    orcid: '0000-0002-6468-120X'
  - name: Paolo Nanni
    affil: 1
    orcid: '0000-0001-8429-3557'
  - name: Jonas Grossmann
    affil: 1; 2
    orcid: '0000-0002-6899-9020'
  - name: Maria D'Errico
    affil: 1; 2
    orcid: '0000-0001-9467-9058'
  - name: Ralph Schlapbach
    affil: 1
    orcid: '0000-0002-7488-4262'
  - name: Christian Panse
    affil: 1; 2
    orcid: '0000-0003-1975-3064'
affiliation:
  - num: 1
    address: Functional Genomics Center Zurich (FGCZ) - Swiss Federal Institute of Technology in Zurich and University of Zurich
  - num: 2
    address: Swiss Institute of Bioinformatics (SIB)
output: 
  bookdown::pdf_document2:
    toc: true
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
abstract: |
  Protein mass spectrometry is widely used for quantitative proteomics
  studies and in particular for relative protein quantification.
  Nevertheless, there is still a need for a flexible and easy-to-use, 
  transparently supporting a variety of well principled statistical
  inference procedures data analysis application programmer interface
  in R. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package integrates
  essential steps of the mass spectrometry-based label-free quantification
  data analysis workflow as: quality control, data normalization, protein
  aggregation, statistical modeling, hypothesis testing, and sample size
  estimation.
  We use `r BiocStyle::Githubpkg("fgcz/prolfqua")` to visualize and model
  simple experimental designs with a single explanatory variable and complex
  experiments with multiple factors.
  Furthermore, we use `r BiocStyle::Githubpkg("fgcz/prolfqua")` to benchmark
  data acquisition, data preprocessing or data modeling methods. Furthermore,
  we apply `r BiocStyle::Githubpkg("fgcz/prolfqua")` to develop highly
  customizable, visually appealing, and interactive data analysis reports
  in PDF or HTML format.
  The R package is available through https://github.com/fgcz/prolfqua.
---

```{r setup, include=FALSE}
## See ACS guidelines [https://publish.acs.org/publish/author_guidelines?coden=ancham#manuscript_text_components]
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 7, fig.height = 5)
```

\newpage

@@

Label-Free Quantification (LFQ) proteomics experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment vs. control are used [TODO: cite]. More recently, more complex experimental designs are implemented with increasing numbers of samples, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [TODO: cite]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a factor or interactions are significant, or differences between factor levels can be estimated, and their significance tested.  Proteins can be ordered or grouped using the estimated fold-change, $t$-statistics, scaled $p$-value, or false discovery rate (FDR) and then subjected to gene set enrichment or over-representation analysis [TODO ...] or validation experiments.

@@

Models in `r BiocStyle::Githubpkg("fgcz/prolfqua")` are specified using R’s linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure [@faraway2016extending,@MASS2002] is an advantage when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis which was performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` using base R or other programming languages by reading the analysis script. However, acknowledging the R formula interface’s complexity to non-statisticians and the popularity of `r BiocStyle::Biocpkg("MSstats")`, we provide the functionality to suggest a model formula from the sample annotation provided in tabular form similarily to `r BiocStyle::Biocpkg("MSstats")` [@Choi2014].

@@

__R__ linear model and linear mixed effect models allow modeling parallel designs, repeated measurements, factorial designs, and many more. __R__'s formula interface for linear models is flexible, widely used, and well documented. This makes it easy to reproduce an analysis performed with _prolfqua_ in other statistical programming languages. We implemented features specific to high throughput experiments, such as the experimental Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We use R6 to model the statistical modeling functionality in `prolfqua` (see Figure \@ref(fig:ContrastUML)).

(ref:ContrastUML) UML diagram of modeling and contrast related classes. Different strategies, e.g., _lm_, _lmer_, and _glm_ can be used to fit various types of models using a model builder (see code). All classes computing contrasts implement the _Contrast_ interface.

@@

\begin{equation} 
s_p^2=\frac{\sum_{i=1}^k (n_i - 1)s_i^2}{\sum_{i=1}^k (n_i - 1)}
\end{equation}

@@

\begin{equation} 
s = \sqrt{\frac{2n_g s_p^2}{n}},
\end{equation}

@@

\begin{equation} 
\tilde{t}_{pj} = \frac{t_{pj} s_p}{\tilde{s}_p}
\end{equation}

@@

To summarize peptide level models to protein models we did applied the method suggested by [@Suomi2017bEnhanced], that is to use the median scaled $p$-value of the peptide models and cumulative distribution function of the beta distribution function to determine a regulation probability of the protein.

To obtain the median $p$-value of the protein, we first rescaled the peptide $p$-values by taking the direction of the fold-change $\hat \beta$ into account, i.e.:

@@

\begin{equation}
\tilde{p} = 1 - |\tilde{p}_{s}|
\end{equation}

Because we used the median as the i-th order statistic  $i = \frac{n}{2} + 0.5$. Therefore, $\gamma = i = \frac{n}{2} + 0.5$ and $\delta = n - i + 1 = n - (\frac{n}{2} + 0.5) + 1 = \frac{n}{2} + 0.5 = \gamma$ are used to parameterize the CDF of the Beta distribution. We implemented this method in the class `ContrastROPECA` (Figure \@ref(fig:ContrastUML)).

@@

(ref:exampleContrasts) Left panel - histogram of $p$-values. Right panel - volcano plots.

@@

The Benchmark functionality of _prolfqua_ includes ROC curves and computes partial areas under those curves (pAUC) and other scores, e.g. the false discovery proportion FDP. We use it to study how well linear, mixed effect models or $p$-value moderation model quantitative mass spectrometric high throughput experiments and compare it with results produced by MSstats [@MSstats2014] and by proDA [@bioxrvproDA2020]. Figure \@ref(fig:FDRfdp)

@@

We can use various statistics to examine the alternative hypothesis $\beta \ne 0$: the contrast estimate, i.e. the $\log_2$ fold-change $\beta$, the $t$-statistic $\frac{\beta}{\sqrt{var(\beta)}}$, or the $p$-value and moderated $p$-value. For each statistic and each value of the statistics we then compute a confusion matrix (see Table \@ref(tab:confusionMatrix)). From the confusion matrix we obtain measures such as true positive rate (TPR), false positive rate (FPR), or false discovery proportion (FDP) which are given by:

@@

\begin{eqnarray}
TPR &= \frac{TP}{TP+FN} = \frac{TP}{P}\\
FPR &= \frac{FP}{FP+TN} = \frac{FP}{N}\\
FDP &= \frac{FP}{TP + FP} = \frac{FP}{R}
\end{eqnarray}

@@

Figure \@ref(fig:FDRfdp) left panel, shows how various estimates obtained from the models, i.e. $\log_2(FC)$, $t$-statistics and scaled $p$-values allow to identify true positives (TP) given a false positive rate. When testing for differential expression, the $\log_2(FC)$ is outperformed by than the $t$-statistic or $p$-value. 
If an accurate estimate of the $log_2(FC)$ is essential, the linear models fitted to protein intensities, estimated using Tukey's median polish, show the best performance. We speculate that this is because of the robustness of Tukey's median polish method. This hypothesis could be examined, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`, into the benchmark.

There are only minor differences in the $p$-AUC when using the $t$-statistics or the scaled $p$-value  (see Figure \@ref(fig:FDRfdp) right panel). In order to compute the confusion matrices based on the $p$-value we first need to rescale it (see Equation \@ref(eq:scalepvalue)). Thus, the $p$-value is a function of the $t$-statistics and the degrees of freedom. Interestingly, the pAUC increases when using $p$-values instead of the $t$-statistics for linear models, while it decreases for mixed effect models. The reason is erroneous denominator degrees of freedom estimation for many proteins in the mixed-effect models.

@@

(ref:FDRfdp) Left panel - Partial area under the ROC curve at $10\%$ FPR for all contrasts and three different statistics: the difference among conditions, the scaled $p$-value (sign(diff) * p.value) and the $t$-statistics (higher is better). The red line average area under the curve. Left panel - Number and percentage of estimated contrasts for each modeling method (higher is better). Right panel - Compare FDR estimate with false discovery proportion (FDP) closer to the diagnoal (red continous line is better).

@@

The probabilistic dropout analysis implemented in the _proDA_ produces inferences comparable to other methods (Figure \@ref(fig:FDRfdp) for the largest number of proteins (Figure \@ref(fig:FDRfdp) left panel). Moreover, the estimated fold changes are unbiased and show high diagnostic accuracy. Furthermore, the scaled $p$-values and $t$-statics are on par with those obtained by the linear model with variance moderation. Therefore, we plan to integrate the _proDA_ package as an additional modeling option into _prolfqua_ (see Figure \@ref(fig:ContrastUML)).

Leveraging on these computer experiments, we can provide the following advice. First, aggregate the peptide intensities to protein intensities using a robust regression method. Second, use linear models as long as they suffice to model your experimental design. Use mixed effect models only if you have to (e.g., repeated measurements). If you use fixed-effect linear models, apply variance moderation since this improves the $t$-statistics and $p$-value estimates. If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the fold change. You can select proteins to study them further based on the FDR, although expect a $bias$.

@@

The authors thank the technology platform fund (TPF) of the University of Zurich.

# Abbreviations

|       |                                                   |
|:------|:--------------------------------------------------|
|AUC    |Area Under the Curve                               |
|ESI-MS |Electro-Spray-Ionization Mass Spectrometry         |
|FDR    |False Discovery Rate                               |
|HPLC   |High-Pressure Liquid Chromatography                |
|LC-MS  |Liquid Chromatography followed by Mass Spectrometry|
|LFQ    |Label-Free Quantification                          |
|LOD    |Limit Of Detection                                 |
|MS     |mass spectrometry                                  |
|OO     |Object-Oriented                                    |

# References {-}

<div id="refs"></div>
